---
title: "batch_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data in

```{r loading}
#load the library & phenotype file
setwd("/mnt/data1/BDR/QC/batch_analysis")
library(methylumi)
library(wateRmelon)
require(gdata)
library(minfi)
library(ggplot2)
require(gridExtra)
library(plyr)
require(IlluminaHumanMethylationEPICmanifest)
library(dplyr)
library(ggfortify)

#Read in the ready made idats objects
setwd("/mnt/data1/BDR/QC/combined")
load("BDR_Mset.rdat")
load("BDR_RGset.rdat")
betas<-betas(msetEPIC)
setwd("/mnt/data1/BDR/QC/batch_analysis")

#load the correct pheno
pheno <- read.csv("/mnt/data1/BDR/QC/combined/pheno_KCL_Bris_Ox1_Ox2_Sex_Braak.csv", header = T,  stringsAsFactors = FALSE)
pheno$Oregon<-ifelse(pheno$Brain_ID=="M1481c2"|pheno$Brain_ID=="M7342c1"|pheno$Brain_ID=="M130c1"|pheno$Brain_ID=="M1205c2", TRUE,FALSE)
pheno$Empty<-pheno$Brain_ID=="EMPTY"
pheno$Control<-pheno$Brain_ID=="Blank"
rownames(pheno) <- pheno$Basename
pheno <- pheno[order(rownames(pheno)),]
betas <- betas[,order(colnames(betas))]
identical(rownames(pheno), colnames(betas))

#form a QC metrics and samples that failed at what steps
QCmetrics<-pheno
SamplesFail<-as.logical(rep("FALSE", nrow(pheno)))
Stepsummary<-as.data.frame(matrix(ncol=0, nrow=2))
rownames(Stepsummary)<-c("Failed This Step", "Total Failed")
```

## Check Signal Intesities

The intensity check is the biggest indicator of sample quality. The median methylated signal intensity and unmethylated signal intensity for each sample is calculcated.

```{r raw intensities, echo=FALSE}
####Check Signal Intesities####
#the median intensities for both signals are calculated and added to the QCmetrics
# m_intensities<-methylated(msetEPIC)
# u_intensities<-unmethylated(msetEPIC)
# M.median<-apply(m_intensities, 2, median)
# U.median<-apply(u_intensities, 2, median)
# QCmetrics<-cbind(pheno,M.median, U.median)

#plot intensities
# coloured by chip
plotfactor<-factor(pheno$SentrixID, levels=c(unique(pheno$SentrixID), "FullyMethylated", "Oregon", "Empty"))
plotfactor[pheno$Control]<-"FullyMethylated"
plotfactor[pheno$Oregon]<-"Oregon"
plotfactor[pheno$Empty]<-"Empty"
par(mfrow = c(1,2))
hist(M.median, xlab = "Median M intensity", main="Histogram of Median Methylated Intensities", cex.main=0.7)
hist(U.median, xlab = "Median U intensity", main="Histogram of Median Unmethylated Intensities", cex.main=0.7)
par(mfrow = c(1,1))
plot(M.median, U.median, pch = 16, xlab = "Median M intensity", ylab = "Median U intensity", col = rainbow(nlevels(plotfactor))[factor(plotfactor)], main="Scatter plot of Signal Intensities coloured by chip")
par(xpd=TRUE)
legend("topright", levels(factor(plotfactor)), col = rainbow(nlevels(plotfactor)), pch = 10, cex=0.5)

#coloured by plate
pheno$Plate=as.character(pheno$Plate)
plotfactor<-factor(pheno$Plate, levels=c(unique(pheno$Plate), "FullyMethylated", "Oregon", "Empty"))
plotfactor[pheno$Control]<-"FullyMethylated"
plotfactor[pheno$Oregon]<-"Oregon"
plotfactor[pheno$Empty]<-"Empty"
par(mfrow = c(1,2))
hist(M.median, xlab = "Median M intensity", main="Histogram of Median Methylated Intensities", cex.main=0.7)
hist(U.median, xlab = "Median U intensity", main="Histogram of Median Unmethylated Intensities", cex.main=0.7)
par(mfrow = c(1,1))
plot(M.median, U.median, pch = 16, xlab = "Median M intensity", ylab = "Median U intensity", col = rainbow(nlevels(plotfactor))[factor(plotfactor)], main="Scatter plot of Signal Intensities coloured by plate")
par(xpd=TRUE)
legend("topright", levels(factor(plotfactor)), col = rainbow(nlevels(plotfactor)), pch = 10, cex=0.5)

```

The plot shows a trend across based on plates. This is due to technical reasons 

```{r PCA of intensities}
#plot pca of intensities
pca_intensity <- as.data.frame(cbind(QCmetrics$M.median, QCmetrics$U.median))
intensity_pca <- prcomp(pca_intensity, center = TRUE, scale. = TRUE)                     
print(summary(intensity_pca))
par(mfrow = c(1, 2))
autoplot(intensity_pca, data = QCmetrics, colour = 'Plate')
autoplot(intensity_pca, data = QCmetrics, colour = 'Institute')
```
We can note here that insititution is causing the most variation. Lets remove those samples with NA 

```{r pca without NAs}
#plot PCA again but without NA, blank or missing 
QCmetrics2 <- na.omit(QCmetrics)
pca_intensity <- as.data.frame(cbind(QCmetrics2$M.median, QCmetrics2$U.median))
intensity_pca <- prcomp(pca_intensity, center = TRUE, scale. = TRUE)                     
summary(intensity_pca)
par(mfrow = c(1, 2))
autoplot(intensity_pca, data = QCmetrics2, colour = 'Institute')
#grouping is unclear - lets see if it can cluster itself
autoplot(intensity_pca, data = QCmetrics2, colour = 'Institute', frame = T) #still a lot of overlap
```


```{r pca by intensities}
pca_intensity <- as.data.frame(cbind(QCmetrics$M.median, QCmetrics$U.median))
intensity_pca <- prcomp(pca_intensity, center = TRUE, scale. = TRUE)                     
summary(intensity_pca)
par(mfrow = c(1, 2))
autoplot(intensity_pca, data = QCmetrics, colour = 'M.median')
autoplot(intensity_pca, data = QCmetrics, colour = 'U.median')
```


##Compare within group variation
We will look at within group varation - most interestingly Oxford as that was most varied on the PCA.
```{r within group variation - Oxford}
###Look into within group variation
#Oxford samples interesting have the most variation for intensitites
QCmetrics_ox <- filter(QCmetrics, QCmetrics$Institute == 'Oxford')
pca_intensity <- as.data.frame(cbind(QCmetrics_ox$M.median, QCmetrics_ox$U.median))
intensity_pca <- prcomp(pca_intensity, center = TRUE, scale. = TRUE)                     
summary(intensity_pca)
par(mfrow = c(3, 2))
autoplot(intensity_pca, data = QCmetrics_ox, colour = 'Plate')
autoplot(intensity_pca, data = QCmetrics_ox, colour = 'Chip')
autoplot(intensity_pca, data = QCmetrics_ox, colour = 'Position')
autoplot(intensity_pca, data = QCmetrics_ox, colour = 'Gender')
autoplot(intensity_pca, data = QCmetrics_ox, colour = 'Age')
```
We can see that the most intensity variation is likely due to plate and chip. We can run an ANOVA within-group statistics to see which variables causes so much variation.

```{r normality test}
QCmetrics_ox$Plate <- as.factor(QCmetrics_ox$Plate) 
QCmetrics_ox$intens_ratio <-as.integer(QCmetrics_ox$M.median/QCmetrics_ox$U.median)
par(mfrow = c(2,2))
hist(QCmetrics_ox$M.median, xlab = "Median M intensity", 
     main="Histogram of normal Median Methylated Intensities", cex.main=0.7)
qqnorm(QCmetrics_ox$M.median,main="QQ plot of normal data(M,median)",pch=19)
qqline(QCmetrics_ox$M.median)


hist(QCmetrics_ox$U.median, xlab = "Median U intensity", 
     main="Histogram of Median Unmethylated Intensities", cex.main=0.7)
qqnorm(QCmetrics_ox$U.median,main="QQ plot of normal data(U.median)",pch=19)
qqline(QCmetrics_ox$U.median)

```




```{r looking at variation within plates}
QCmetrics_ox$Plate <- as.factor(QCmetrics_ox$Plate) 
par(mfrow = c(1, 2))
ggplot(QCmetrics_ox, aes(QCmetrics_ox$Plate, QCmetrics_ox$M.median)) +
  geom_violin(fill = "lightblue3") +
  geom_boxplot() +
  labs(x = "Plates", y = "M.median intensity ",
        title ="Methylated median intesity across plates") 

ggplot(QCmetrics_ox, aes(QCmetrics_ox$Plate, QCmetrics_ox$U.median)) +
  geom_violin(fill = "thistle1") +
  geom_boxplot()+
  labs(x = "Plates", y = "U.median intensity ",
       title ="Unmethylated median intesity across plates")

```
```{r ANOVA and Post-Hoc test}
QCmetrics_ox$intens_ratio <-as.integer(QCmetrics_ox$M.median/QCmetrics_ox$U.median)
anova1 <-aov(QCmetrics_ox$intens_ratio~QCmetrics_ox$Plate)
print(summary(anova1))
pht <- TukeyHSD(anova1)
print(pht)
pht1 <- as.data.frame(pht$`QCmetrics_ox$Plate`)
p_adj <- pht1$`p adj` <- as.factor(pht1$`p adj`)
ggplot(pht1, aes(rownames(pht1), pht1$diff, fill = p_adj)) +
  geom_col() +
  ylim(-0.05, 0.05) +
  labs(x = "Plates", y = "Mean Difference",
       title ="Mean differences between Plates") +
  theme(axis.text.x = element_text(angle = 45))
```





